---
title: "Chapter 5: Data Transformation"
output: html_notebook
---

```{r}
library(nycflights13)
library(tidyverse)
```

# 5.2 Filter rows with `filter()`

## 5.2.4 Exercises

1. Find all flights that:

a. Had an arrival delay of two or more hours

```{r}
filter(flights, arr_delay >= 120)
```

b. Flew to Houston (IAH or HOU)

```{r}
filter(flights, dest %in% c("IAH", "HOU"))
```

c. Were operated by United, American, or Delta

```{r}
airlines
filter(flights, carrier %in% c("UA", "AA", "DL"))
```

d. Departed in summer (July, August, and September)

```{r}
filter(flights, month %in% c(7, 8, 9))
```

e. Arrived more than two hours late, but didn’t leave late

```{r}
filter(flights, arr_delay > 120, dep_delay <= 0)
```

f. Were delayed by at least an hour, but made up over 30 minutes in flight

```{r}
filter(flights, arr_delay >= 60, arr_delay - dep_delay < -30)
```

g. Departed between midnight and 6am (inclusive)

```{r}
filter(flights, dep_time <= 600 | dep_time == 2400)
```

2. Another useful dplyr filtering helper is `between()`. What does it do? Can you use it to simplify the code needed to answer the previous challenges?

**`between()` returns items that arehave a variable value between two boundary values (inclusive, ie it tests for `>=` and `<=` on the left and right boundaries).**

d. Departed in summer (July, August, and September)

```{r}
filter(flights, between(month, 7, 9))
```

3. How many flights have a missing `dep_time`? What other variables are missing? What might these rows represent?

```{r}
filter(flights, is.na(dep_time))
```

**These rows are also missing `dep_delay`, `arr_time`, `arr_delay` and `air_time`. Given that they have all the scheduled details but are missing all actual flight data, these rows appear to represent cancelled flights.**

4. Why is `NA ^ 0` not missing? Why is `NA | TRUE` not missing? Why is `FALSE & NA` not missing? Can you figure out the general rule? (`NA * 0` is a tricky counterexample!)

```{r}
NA ^ 0
NA | TRUE
NA | FALSE # Counter-example
FALSE & NA
TRUE & NA # Counter-example
NA * 0
```

**`NA ^ 0` is not missing because any value to the power of zero equals 1 (although I don't have an explanation right now why the same principle doesn't apply for `NA * 0`). `NA | TRUE` is not missing because only one side of the 'or' operator needs to evaluate as true (conversely, note that `NA | FALSE` is missing). `FALSE & NA` is not missing because both sides of the 'and' operator would need to evaluate TRUE for it to be true so based on the right-hand side it would be false regardless of the actual value of the `NA` (conversely, note that `TRUE & NA` is missing).**

# 5.3 Arrange rows with `arrange()`

## 5.3.1 Exercises

1. How could you use `arrange()` to sort all missing values to the start? (Hint: use `is.na()`).

**Using `dep_time` as an example.**

```{r}
arrange(flights, desc(is.na(dep_time)))
```

2. Sort flights to find the most delayed flights. Find the flights that left earliest.

```{r}
arrange(flights, desc(arr_delay), dep_delay)
```

3. Sort flights to find the fastest flights.

```{r}
arrange(flights, air_time)
```

4. Which flights travelled the longest? Which travelled the shortest?

```{r}
arrange(flights, desc(distance))
```

```{r}
arrange(flights, distance)
```

# 5.4 Select columns with `select()`

## 5.4.1 Exercises

1. Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay` from flights.

```{r}
select(flights, dep_time, dep_delay, arr_time, arr_delay)
```

```{r}
select(flights, 4, 6, 7, 9)
```

```{r}
select(flights, starts_with("dep_"), starts_with("arr_"))
```

**Could keep going with a minus operator to drop all of the other columns, etc.**

2. What happens if you include the name of a variable multiple times in a `select()` call?

```{r}
select(flights, dep_time, dep_time)
```

**It doesn't duplicate the variable.**

3. What does the `one_of()` function do? Why might it be helpful in conjunction with this vector?

**It allows selection of variables by matching against a vector of strongs. In the code below I've used it to select all of the variables that _aren't_ listed in the vector.**

```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
select(flights, -one_of(vars))
```

4. Does the result of running the following code surprise you? How do the `select` helpers deal with case by default? How can you change that default?

```{r}
select(flights, contains("TIME"))
```

**By default the `select` helpers are case-insensitive. It can be modified by passing the argument `ignore.case = FALSE`.**

```{r}
select(flights, contains("TIME", ignore.case = FALSE))
```
