---
title: "Chapter 7: Exploratory Data Analysis"
output: html_notebook
---

```{r}
library(tidyverse)
```

# 7.3 Variation

## 7.3.4 Exercises

1. Explore the distribution of each of the `x`, `y`, and `z` variables in diamonds. What do you learn? Think about a diamond and how you might decide which dimension is the length, width, and depth.

```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = x), binwidth = 0.5) +
  coord_cartesian(xlim = c(0, 12), ylim = c(0, 30000))
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
  coord_cartesian(xlim = c(0, 12), ylim = c(0, 30000))
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = z), binwidth = 0.5) +
  coord_cartesian(xlim = c(0, 12), ylim = c(0, 30000))
```

**`x` and `y` are similarly distributed in terms of values, while `z` tends to have lower values and less variability than the other two dimensions. When measuring you would probably treat the longest dimensions as width and height and then the third dimension would become depth and be measured last.**

2. Explore the distribution of `price`. Do you discover anything unusual or surprising? (Hint: Carefully think about the `binwidth` and make sure you try a wide range of values.)

```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = price))
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = price),  binwidth = 500)
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = price),  binwidth = 100)
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = price),  binwidth = 50)
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = price),  binwidth = 10)
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = price),  binwidth = 10) +
  coord_cartesian(xlim = c(1000, 2000))
```

**The general distribution shows that the most common prices are relatively low (most common price is around \$1,000), and higher-priced diamonds become increasingly rare. But there is also an unusual part of the distribution around \$1,500 where no diamonds fall into that price range.**

3. How many diamonds are 0.99 carat? How many are 1 carat? What do you think is the cause of the difference?

```{r}
diamonds %>%
  filter(between(carat, 0.99, 1.0)) %>%
  count(carat)
```

**There are many more 1 carat diamonds. This could just be rounding or lack of precision in measuring, but it seems likely that it would be more appealing to classify diamonds as 1 carat because they would be more marketable.**

4. Compare and contrast `coord_cartesian()` vs `xlim()` or `ylim()` when zooming in on a histogram. What happens if you leave binwidth unset? What happens if you try and zoom so only half a bar shows?

```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = price),  binwidth = 10) +
  coord_cartesian(xlim = c(1000, 2000))
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = price),  binwidth = 10) +
  xlim(1000, 2000)
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = price)) +
  coord_cartesian(xlim = c(1000, 2000))
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = price)) +
  xlim(1000, 2000)
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = price),  binwidth = 100) +
  coord_cartesian(xlim = c(1000, 1050))
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = price),  binwidth = 100) +
  xlim(1000, 1050)
```

**Note that `xlim()` removes rows from the dataset that is being plotted, while `coord_cartesian()` keeps the full dataset but then only plots a limited range. If you leave `binwidth` unset then this means `coord_cartesian` will use default binwidths based on the full range of data but `xlim()` will break up the range that has been left in the data. And if you try to zoom so that only half a bar shows then `coord_artesian` will display the half-bar but `xlim()` doesn't because it no longer has the correct range of data to plot that bar.**

# 7.4 Missing values

## 7.4.1 Exercises

```{r}
diamonds2 <- diamonds %>% 
  mutate(y = ifelse(y < 3 | y > 20, NA, y))
```

1. What happens to missing values in a histogram? What happens to missing values in a bar chart? Why is there a difference?

```{r}
ggplot(data = diamonds2) +
  geom_histogram(mapping = aes(x = y))
```

```{r}
ggplot(data = diamonds2) +
  geom_bar(mapping = aes(x = y))
```

**This needs more examination. Might come back to it.**

2. What does `na.rm = TRUE` do in `mean()` and `sum()`?

```{r}
diamonds2 %>%
  summarise(mean(y), sum(y))
```

```{r}
diamonds2 %>%
  summarise(mean(y, na.rm = TRUE), sum(y, na.rm = TRUE))
```

**`na.rm = TRUE` removes missing values and calculates means and sums based on the set of non-missing values in the variable.**

# 7.5 Covariation

## 7.5.1 A categorical and continuous variable

### 7.5.1.1 Exercises

1. Use what you’ve learned to improve the visualisation of the departure times of cancelled vs. non-cancelled flights.

```{r}
nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>%
  ggplot(mapping = aes(x = sched_dep_time, y = ..density..)) +
  geom_freqpoly(aes(colour = cancelled))
  
```

**This suggests that the proportion of flights cancelled tends to increase later in the day, especially with scheduled departure times from 3pm onward.**

2. What variable in the diamonds dataset is most important for predicting the price of a diamond? How is that variable correlated with cut? Why does the combination of those two relationships lead to lower quality diamonds being more expensive?

```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price))
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = cut, y = price))
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = color, y = price))
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = clarity, y = price))
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = x, y = price))
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = y, y = price))
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = z, y = price))
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = depth, y = price))
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = table, y = price))
```

**Higher carat diamonds tend to hace higher prices.**

```{r}
ggplot(data = diamonds) +
  geom_boxplot(aes(x = cut, y = carat))
```

**Higher carat diamonds are more common in the lower-quality classes of cut, which appears to explain why low-quality cut diamonds tend to have higher prices.**

3. Install the ggstance package, and create a horizontal boxplot. How does this compare to using `coord_flip()`?

```{r}
library(ggstance)
ggplot(data = mpg) +
  geom_boxploth(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy))
ggplot(data = mpg) +
  geom_boxploth(mapping = aes(y = reorder(class, hwy, FUN = median), x = hwy))
```

**The key difference is that with ggstance you assign to the `y` axis the variable that would be assigned to the `x` axis before applying `coord_flip()`, and vice versa. In other words, ggstance maps directly onto the axes as they will appear in the horizontal plot.**

4. One problem with boxplots is that they were developed in an era of much smaller datasets and tend to display a prohibitively large number of “outlying values”. One approach to remedy this problem is the letter value plot. Install the lvplot package, and try using `geom_lv()` to display the distribution of price vs cut. What do you learn? How do you interpret the plots?

```{r}
library(lvplot)
ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
  geom_lv()
```

**There appears to be a package conflict that will need to be resolved.**

5. Compare and contrast `geom_violin()` with a facetted `geom_histogram()`, or a coloured `geom_freqpoly()`. What are the pros and cons of each method?

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
  geom_violin()
ggplot(data = diamonds, mapping = aes(x = price)) +
  geom_histogram() +
  facet_wrap(~ cut)
ggplot(data = diamonds, mapping = aes(x = price)) +
  geom_freqpoly(mapping = aes(colour = cut))
```

6. If you have a small dataset, it’s sometimes useful to use `geom_jitter()` to see the relationship between a continuous and categorical variable. The ggbeeswarm package provides a number of methods similar to `geom_jitter()`. List them and briefly describe what each one does.

```{r}
library(ggbeeswarm)
?ggbeeswarm
```

**Look at `position_quasirandom()` and `position_beeswarm()`.**

## 7.5.2 Two categorical variables

### 7.5.2.1 Exercises

1. How could you rescale the count dataset above to more clearly show the distribution of cut within colour, or colour within cut?

```{r}
diamonds %>%
  group_by(cut, color) %>%
  summarise(n = n()) %>%
  mutate(prop = n/sum(n)) %>%
  ggplot(mapping = aes(x = color, y = cut)) +
    geom_tile(mapping = aes(fill = prop))
```

**Calculating the proportion of colours within each cut rather than frequencies allows us to see the distribution of colours within each cut level more clearly.**

2. Use `geom_tile()` together with dplyr to explore how average flight delays vary by destination and month of year. What makes the plot difficult to read? How could you improve it?

```{r}
library(nycflights13)
flights %>%
  group_by(dest, month) %>%
  summarise(avg_delay = mean(arr_delay, na.rm = TRUE)) %>%
  ggplot(mapping = aes(x = as.factor(month), y = dest)) +
    geom_tile(mapping = aes(fill = avg_delay))
```

**The number of destinations makes the y-axis overcrowded and even with the chart opened in a full-screen window it isn't clearly readable.**

3. Why is it slightly better to use `aes(x = color, y = cut)` rather than `aes(x = cut, y = color)` in the example above?

```{r}
ggplot(data = diamonds) +
  geom_count(mapping = aes(x = cut, y = color))
ggplot(data = diamonds) +
  geom_count(mapping = aes(x = color, y = cut))
```

**Having colour on the x-axis narrows the width of the chart which seems to make comparison of points easier.**


## 7.5.3 Two continuous variables

### 7.5.3.1 Exercises

1. Instead of summarising the conditional distribution with a boxplot, you could use a frequency polygon. What do you need to consider when using `cut_width()` vs `cut_number()`? How does that impact a visualisation of the 2d distribution of carat and price?

```{r}
smaller <- diamonds %>% 
  filter(carat < 3)
```

**I'll need to come back to this.**

2. Visualise the distribution of carat, partitioned by price.

```{r}
ggplot(data = smaller, mapping = aes(x = price, y = carat)) +
  geom_boxplot(mapping = aes(group = cut_width(price, 1000)), varwidth = TRUE)
```

3. How does the price distribution of very large diamonds compare to small diamonds. Is it as you expect, or does it surprise you?

```{r}
ggplot(data = smaller, mapping = aes(x = carat, y = price)) +
  geom_boxplot(mapping = aes(group = cut_width(carat, 0.1)))
```

**The general trend that larger diamonds have higher average prices is a basic finding I would have expected, but the distributino boxplots show some other interesting patterns. The width of distributions is very narrow for the smallest diamonds so they are all very low price, but distributions become wider as the size increases, so it's far from the case that all larger diamonds are higher-priced than some smaller diamonds. It's around the 1 carat mark that the distribution becomes wider so that some outliers at that size have very high prices, but by the time you reach 2 carats those high prices have become the norm but there are now low-priced outliers instead.**

4. Combine two of the techniques you’ve learned to visualise the combined distribution of cut, carat, and price.

```{r}
ggplot(data = smaller, mapping = aes(x = carat, y = price)) +
  geom_boxplot(mapping = aes(group = cut_width(carat, 0.1))) +
  facet_grid(cut ~ .)
```

5. Two dimensional plots reveal outliers that are not visible in one dimensional plots. For example, some points in the plot below have an unusual combination of x and y values, which makes the points outliers even though their x and y values appear normal when examined separately.

```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = x, y = y)) +
  coord_cartesian(xlim = c(4, 11), ylim = c(4, 11))
```

Why is a scatterplot a better display than a binned plot for this case?

```{r}
ggplot(data = diamonds) +
  geom_bin2d(mapping = aes(x = x, y = y)) +
  coord_cartesian(xlim = c(4, 11), ylim = c(4, 11))
```

```{r}
ggplot(data = diamonds) +
  geom_boxplot(mapping = aes(x = x, y = y, group = cut_width(x, 0.5))) +
  coord_cartesian(xlim = c(4, 11), ylim = c(4, 11))
```

**The relationship is so strong and most points tightly clustered so they don't break into useful subgroupings when binned.**
