---
title: "Chapter 11: Data import"
output: html_notebook
---

```{r}
library(tidyverse)
```

# 11.2 Getting started

## 11.2.2 Exercises

1. What function would you use to read a file where fields were separated with “|”?

```{r}
read_delim(
  "a|b|c
  1|2|3
  4|5|6",
  "|"
)
```

**`read_delim` can be used to read in files with an arbitrary character as the delimiter.**

2. Apart from file, skip, and comment, what other arguments do read_csv() and read_tsv() have in common?

```{r}
?read_csv
```

read_csv(file, col_names = TRUE, col_types = NULL,
  locale = default_locale(), na = c("", "NA"), quoted_na = TRUE,
  comment = "", trim_ws = TRUE, skip = 0, n_max = Inf,
  guess_max = min(1000, n_max), progress = interactive())

read_tsv(file, col_names = TRUE, col_types = NULL,
  locale = default_locale(), na = c("", "NA"), quoted_na = TRUE,
  comment = "", trim_ws = TRUE, skip = 0, n_max = Inf,
  guess_max = min(1000, n_max), progress = interactive())

**Other arguments are `col_names`, `col_types`, `locale`, `na`, `quoted_na`, `trim_ws`, `n_max`, `guess_max`, and `progress`.**

3. What are the most important arguments to read_fwf()?

```{r}
?read_fwf
```
read_fwf(file, col_positions, col_types = NULL, locale = default_locale(),
  na = c("", "NA"), comment = "", skip = 0, n_max = Inf,
  guess_max = min(n_max, 1000), progress = interactive())

**The key arguments are `file` to indicate which file to read and `fwf_positions` or `fwf_width` (which provide the required info for `col_positions`), and then there are options to set `na`, `col_types` and `skip` to suit the data file, along with some other arguments that may be used occasionally.**

4. Sometimes strings in a CSV file contain commas. To prevent them from causing problems they need to be surrounded by a quoting character, like " or '. By convention, read_csv() assumes that the quoting character will be ", and if you want to change it you’ll need to use read_delim() instead. What arguments do you need to specify to read the following text into a data frame?

`"x,y\n1,'a,b'"`

```{r}
read_delim(
  "x,y\n1,'a,b'",
  delim = ",",
  quote = "\'"
)
```

5. Identify what is wrong with each of the following inline CSV files. What happens when you run the code?

``` {r}
read_csv("a,b\n1,2,3\n4,5,6")
```

**There are only two variables in the header row but three values in each subsequent row - the final values are dropped as the tibble is only set up with two variables.**

```{r}
read_csv("a,b,c\n1,2\n1,2,3,4")
```

**There are three variables defined in the header row but the data rows have either one value too few or one too many. The excess value is ignored, while when there aren't enough values in a row the final column is `NA`.**

```{r}
read_csv("a,b\n\"1")
```

**Again, it's one value short so ends up with an `NA`. It also has an escaped quotation mark before the value but is missing the closing escaped quotation mark after it, but the function still seems to read the value in okay.**

```{r}
read_csv("a,b\n1,2\na,b")
```

**There doesn't seem to be anything absolutely wrong about this file, but note that the variables become character because of the second row of data. If those values weren't there then the first row would be read in as numeric variables.**

```{r}
read_csv("a;b\n1;3")
```

**It looks like this should have used `read_csv2` for the semicolons to be considered as separators - see below.**

```{r}
read_csv2("a;b\n1;3")
```

# 11.3 Parsing a vector

## 11.3.5 Exercises

1. What are the most important arguments to `locale()`?

```{r}
?locale
```

**`decimal_mark` and `grouping_mark` are important for processing of all numeric data. `tz` will be important for any time data to ensure it accurately reflects the local time, and then `date_names`, `date_format` and `time_format` may also be important for processing date and time values. `encoding` is crucial and will be problematic if set wrong for the input file, but in many modern cases the default of UTF-8 will be correct and covers everything.**

2. What happens if you try and set `decimal_mark` and `grouping_mark` to the same character? What happens to the default value of `grouping_mark` when you set `decimal_mark` to “,”? What happens to the default value of `decimal_mark` when you set the `grouping_mark` to “.”?

```{r}
locale(decimal_mark = ".", grouping_mark = ".")
```

**If the two arguments try to use the same character an error is thrown - there must be different marks for decimal vs grouping.**

```{r}
locale(decimal_mark = ",")
```

**The grouping mark becomes `.`.**

```{r}
locale(grouping_mark = ".")
```

**The decimal mark becomes `,`.**

3. I didn’t discuss the `date_format` and `time_format` options to `locale()`. What do they do? Construct an example that shows when they might be useful.

```{r}
library(hms)
parse_date("08/09/2010", locale = locale(date_format = "%d/%m/%Y"))
parse_time("1820", locale = locale(time_format = "%H%M"))
```

**These options allow you to define the way dates and times are formatted so they can be parsed correctly.**

4. If you live outside the US, create a new locale object that encapsulates the settings for the types of file you read most commonly.

```{r}
locale(
  date_names = "en",
  date_format = "%d/%m/%Y",
  time_format = "%h:%M%p %Z",
  decimal_mark = ".",
  grouping_mark = ",",
  tz = "Australia/Sydney",
  encoding = "UTF-8"
)
```

5. What’s the difference between `read_csv()` and `read_csv2()`?

**The difference is the delimiting character - `,` in the former, `;` in the latter.**

6. What are the most common encodings used in Europe? What are the most common encodings used in Asia? Do some googling to find out.

**Aside from UTF-8, some of the most commonly used encodings in modern documents (ASCII has become less common in the 21st century) would appear to include the ISO-8859 variants (e.g., ISO-8859-1 in Western Europe), GB18030 in China which is a transformation of Unicode, and the Windows character sets (e.g., Windows-1251 for Cyrillic alphabets and Windows-1252 for Western/Latin). Some stats show that [the vast majority of websites are now using UTF-8](https://w3techs.com/technologies/overview/character_encoding/all) but data files may be a different story.**

7. Generate the correct format string to parse each of the following dates and times:

```{r}
d1 <- "January 1, 2010"
d2 <- "2015-Mar-07"
d3 <- "06-Jun-2017"
d4 <- c("August 19 (2015)", "July 1 (2015)")
d5 <- "12/30/14" # Dec 30, 2014
t1 <- "1705"
t2 <- "11:15:10.12 PM"
parse_date(d1, "%B %d, %Y")
parse_date(d2, "%Y-%b-%d")
parse_date(d3, "%d-%b-%Y")
parse_date(d4, "%B %d (%Y)")
parse_date(d5, "%m/%d/%y")
parse_time(t1, "%H%M")
parse_time(t2, "%I:%M:%OS %p")
```
