---
title: "Chapter 13: Relational data"
output: html_notebook
---

# 13.2 `nycflights13`

## 13.2.1 Exercises

```{r}
library(tidyverse)
library(nycflights13)
```

1. Imagine you wanted to draw (approximately) the route each plane flies from its origin to its destination. What variables would you need? What tables would you need to combine?

**You would need `origin` and `dest` for each flight, along with the `lat` and `long` for each origin and destination airport. This means that you would need to combine the `flights` and `airports` tables.**

2. I forgot to draw the relationship between `weather` and `airports`. What is the relationship and how should it appear in the diagram?

**`origin` in `weather` connects to `faa` in `airports`.**

3. `weather` only contains information for the origin (NYC) airports. If it contained weather records for all airports in the USA, what additional relation would it define with flights?

**It would connect with `dest` in `flights`so that weather records for the destination airport could be identified as well.**

4. We know that some days of the year are “special”, and fewer people than usual fly on them. How might you represent that data as a data frame? What would be the primary keys of that table? How would it connect to the existing tables?

**You could have a `specialdays` table with a `dayname` as its primary key and then connect to `flights` via `year`, `month` and `day` variables.**

# 13.3 Keys

## 13.3.1 Exercises

1. Add a surrogate key to `flights`.

```{r}
flights <- flights %>%
  mutate(flight_id = row_number())
flights %>%
  count(flight_id) %>%
  filter(n > 1)
```

2. Identify the keys in the following datasets

  1. `Lahman::Batting`,
  2. `babynames::babynames`
  3. `nasaweather::atmos`
  4. `fueleconomy::vehicles`
  5. `ggplot2::diamonds`

(You might need to install some packages and read some documentation.)

```{r}
library(Lahman)
Batting
```

```{r}
Batting %>%
  count(playerID, yearID, stint) %>%
  filter(n > 1)
```

```{r}
library(babynames)
babynames
```

```{r}
babynames %>%
  count(year, sex, name) %>%
  filter(nn > 1)
```

```{r}
library(nasaweather)
atmos
```

```{r}
atmos %>%
  count(lat, long, year, month) %>%
  filter(n > 1)
```

```{r}
library(fueleconomy)
vehicles
```

```{r}
vehicles %>%
  count(id) %>%
  filter(n > 1)
```

```{r}
diamonds
```

**This doesn't appear to have a primary key because each combination of diamond attributes wouldn't be guaranteed to be unique (even if they are unique within the current data set). It would be best to create a surrogate key.**

```{r}
diamonds <- diamonds %>%
  mutate(id = row_number())
```

3. Draw a diagram illustrating the connections between the `Batting`, `Master`, and `Salaries` tables in the Lahman package. Draw another diagram that shows the relationship between `Master`, `Managers`, `AwardsManagers`.

**`Batting$playerID` connects to `Master$playerID`. `Salaries$playerID` connects to `Master$playerID`. `Batting` and `Salaries` could be connected via `playerID`, `yearID`, and `teamID`.**

**`Master$playerID` connects to `Managers$playerID`. `Master$playerID` connects to `AwardsManagers$playerID`. `Managers` and `AwardsManagers` via `playerID` and `yearID`.**

How would you characterise the relationship between the `Batting`, `Pitching`, and `Fielding` tables?

```{r}
Batting
Pitching
Fielding
```

**They should have a one-to-one relationship with matching primary keys (`playerID`, `yearID`, `stint`) across the three tables - with most position players only having records for `Batting` and `Fielding`, while pitchers are likely to have records only in `Pitching` and `Fielding`.**

# 13.4 Mutating joins

## 13.4.6 Exercises

1. Compute the average delay by destination, then join on the `airports` data frame so you can show the spatial distribution of delays. Here’s an easy way to draw a map of the United States:

```{r}
airports %>%
  semi_join(flights, c("faa" = "dest")) %>%
  ggplot(aes(lon, lat)) +
    borders("state") +
    geom_point() +
    coord_quickmap()
```
(Don’t worry if you don’t understand what `semi_join()` does — you’ll learn about it next.)

You might want to use the `size` or `colour` of the points to display the average delay for each airport.

```{r}
flights %>%
  group_by(dest) %>%
  summarise(avg_delay = mean(arr_delay, na.rm = TRUE)) %>%
  left_join(airports, c("dest" = "faa")) %>%
  ggplot(aes(lon, lat)) +
    borders("state") +
    geom_point(aes(colour = avg_delay)) +
    coord_quickmap()
```

2. Add the location of the origin and destination (i.e. the `lat` and `lon`) to `flights`.

```{r}
flights %>%
  left_join(airports, by = c("dest" = "faa")) %>%
  left_join(airports, by = c("origin" = "faa"))
```

3. Is there a relationship between the age of a plane and its delays?

```{r}
flights %>%
  group_by(tailnum) %>%
  summarise(avg_delay = mean(arr_delay, na.rm = TRUE)) %>%
  right_join(planes) %>%
  ggplot(aes(year, avg_delay)) +
    geom_point()
```

**There doesn't appear to be a clear relationship.**

4. What weather conditions make it more likely to see a delay?

```{r}
flights %>%
  group_by(year, month, day, hour) %>%
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE)) %>%
  left_join(weather) %>%
  ggplot(aes(temp, avg_delay)) +
    geom_point(alpha = 0.1)
flights %>%
  group_by(year, month, day, hour) %>%
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE)) %>%
  left_join(weather) %>%
  ggplot(aes(wind_speed, avg_delay)) +
    geom_point(alpha = 0.1)
flights %>%
  group_by(year, month, day, hour) %>%
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE)) %>%
  left_join(weather) %>%
  ggplot(aes(precip, avg_delay)) +
    geom_point(alpha = 0.1)
flights %>%
  group_by(year, month, day, hour) %>%
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE)) %>%
  left_join(weather) %>%
  ggplot(aes(pressure, avg_delay)) +
    geom_point(alpha = 0.1)
flights %>%
  group_by(year, month, day, hour) %>%
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE)) %>%
  left_join(weather) %>%
  ggplot(aes(visib, avg_delay)) +
    geom_point(alpha = 0.1)
```

5. What happened on June 13 2013? Display the spatial pattern of delays, and then use Google to cross-reference with the weather.

```{r}
flights %>%
  filter(year == 2013, month == 6, day == 13) %>%
  group_by(dest) %>%
  summarise(avg_delay = mean(arr_delay, na.rm = TRUE)) %>%
  left_join(airports, c("dest" = "faa")) %>%
  ggplot(aes(lon, lat)) +
    borders("state") +
    geom_point(aes(colour = avg_delay)) +
    coord_quickmap()
```

**There were severe storms in the mid-Atlantic region.**

# 13.5 Filtering joins

## 13.5.1 Exercises

1. What does it mean for a flight to have a missing `tailnum`? What do the tail numbers that don’t have a matching record in planes have in common? (Hint: one variable explains ~90% of the problems.)

```{r}
flights %>%
  anti_join(planes, by = "tailnum") %>%
  count(carrier)
?planes
```

**A missing `tailnum` means that the metadata for that plane hasn't been recorded. The vast majority of these cases come from two carriers (AA and MQ). Looking at the help file for `planes` explains that these two carriers report fleet numbers rather than tail numbers.**

2. Filter flights to only show flights with planes that have flown at least 100 flights.

```{r}
flights_100 <- flights %>%
  count(tailnum) %>%
  filter(n >= 100)
semi_join(flights, flights_100)
```

3. Combine `fueleconomy::vehicles` and `fueleconomy::common` to find only the records for the most common models.

```{r}
semi_join(vehicles, common)
```

4. Find the 48 hours (over the course of the whole year) that have the worst delays. Cross-reference it with the weather data. Can you see any patterns?

```{r}
worst_delays <- flights %>%
  group_by(year, month, day, hour) %>%
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE)) %>%
  arrange(desc(avg_delay)) %>%
  head(48)
semi_join(weather, worst_delays) %>%
  ggplot(aes(temp)) +
    geom_histogram()
```

**It's possible to plot histograms for the different weather variables in a search for patterns. The temperature example above suggests that sometimes the delays happened on quite cold days but more of them happened in mild or warm weather.**

5. What does `anti_join(flights, airports, by = c("dest" = "faa"))` tell you? What does `anti_join(airports, flights, by = c("faa" = "dest"))` tell you?

```{r}
anti_join(flights, airports, by = c("dest" = "faa"))
```

**This shows flights that went to a destination not listed in the airports database - from a quick inspection and search, many of these seem to be to airports in Puerto Rico or elsewhere in the Caribbean, ie outside the mainland United States.**

```{r}
anti_join(airports, flights, by = c("faa" = "dest"))
```

**This shows airports that are listed in the airports database but which weren't the destination for any flights in the flights database.**

6. You might expect that there’s an implicit relationship between plane and airline, because each plane is flown by a single airline. Confirm or reject this hypothesis using the tools you’ve learned above.

```{r}
# Identify any planes (by `tailnum` that didn't have only a single carrier recorded for their flight records)
multi_carrier <- flights %>%
  count(tailnum, carrier) %>%
  count(tailnum) %>%
  filter(nn != 1)
semi_join(flights, multi_carrier) %>%
  count(tailnum, carrier)
```

