---
title: "Chapter 19: Functions"
output: html_notebook
---

```{r}
library(lubridate)
library(stringr)
```

# 19.2 When should you write a function?

## 19.2.1 Practice

1. Why is `TRUE` not a parameter to `rescale01()`? What would happen if `x` contained a single missing value, and `na.rm` was `FALSE`?

```{r}
rescale01_1 <- function(x) {
  rng <- range(x, na.rm = FALSE)
  (x - rng[1]) / (rng[2] - rng[1])
}
rescale01_1(c(1, 2, 3, NA, 5))
```

**With `na.rm = FALSE` the function doesn't rescale the non-missing values. It seems clear that nobody would want to have the option to produce this behaviour and that in any vector where there are missing values, the scaling should be carried out for all of the non-missing values. For that reason, there is no justification for including a parameter to control that behaviour.**

2. In the second variant of `rescale01()`, infinite values are left unchanged. Rewrite `rescale01()` so that `-Inf` is mapped to `0`, and `Inf` is mapped to `1`.

```{r}
rescale01_2 <- function(x) {
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  x[x == -Inf] <- rng[1]
  x[x == Inf] <- rng[2]
  (x - rng[1]) / (rng[2] - rng[1])
}
rescale01_2(c(1, 2, 3, NA, 5, Inf, -Inf))
```

3. Practice turning the following code snippets into functions. Think about what each function does. What would you call it? How many arguments does it need? Can you rewrite it to be more expressive or less duplicative?

`mean(is.na(x))`

```{r}
# Calculate the proportion of the values in a vector that are missing.
prop_na <- function(x) {
  mean(is.na(x))
}
prop_na(c(1, 2, 3, NA, 5))
```

`x / sum(x, na.rm = TRUE)`

```{r}
# Assuming all non-negative values, calculate values as proportions of the total
prop_of_total <- function(x) {
  x / sum(x, na.rm = TRUE)
}
prop_of_total(c(1, 2, 3, NA, 4))
```

`sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE)`

```{r}
# Coefficient of variation (also known as relative standard deviation)
cv <- function(x) {
  sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE)
}
cv(c(1, 2, 3, NA, 4))
```

4. Follow http://nicercode.github.io/intro/writing-functions.html to write your own functions to compute the variance and skew of a numeric vector.

```{r}
variance <- function(x) {
  m <- mean(x)
  n <- length(x)
  sum((x - m) ^ 2) / (n - 1)
}
var(c(0:10))
variance(c(0:10))
```

```{r}
skewness <- function(x) {
  m <- mean(x)
  v <- var(x)
  n <- length(x)
  third_moment <- sum((x - m) ^ 3) / (n - 2)
  third_moment / (v ^ (3 / 2))
}
normdist <- rnorm(10000)
skewness(normdist)
```

5. Write `both_na()`, a function that takes two vectors of the same length and returns the number of positions that have an `NA` in both vectors.

```{r}
both_na <- function(x, y) {
  sum(is.na(x) & is.na(y))
}
both_na(c(1, 2, 3, NA, 5), c(NA, 2, 3, NA, 5))
both_na(c(1, NA, NA, NA, 5), c(1, NA, 3, NA, 5))
```

6. What do the following functions do? Why are they useful even though they are so short?

`is_directory <- function(x) file.info(x)$isdir`
`is_readable <- function(x) file.access(x, 4) == 0`

**`is_directory` checks whether the value(s) passed to it are directories. `is_readable` checks whether the value(s) passed to it are files/directories that have read permission. The function names make obvious what they do and provide an intuitive way for people to run these checks without having to remember exactly what syntax needs to be used.**

7. Read the [complete lyrics](https://en.wikipedia.org/wiki/Little_Bunny_Foo_Foo) to “Little Bunny Foo Foo”. There’s a lot of duplication in this song. Extend the initial piping example to recreate the complete song, and use functions to reduce the duplication.

**I'll come back to this.**

# 19.3 Functions are for humans and computers

## 19.3.1 Exercises

1. Read the source code for each of the following three functions, puzzle out what they do, and then brainstorm better names.

```{r}
f1 <- function(string, prefix) {
  substr(string, 1, nchar(prefix)) == prefix
}
f2 <- function(x) {
  if (length(x) <= 1) return(NULL)
  x[-length(x)]
}
f3 <- function(x, y) {
  rep(y, length.out = length(x))
}

f1("this is a string", "this")
f1("this is a string", "is a")

f2(c("a", "b", "c", "d"))
f2(1:10)

f3(c(1:5), c("a", "b", "c", "d", "e", "f", "g", "h", "i", "j"))
f3(c(1:5), c("a", "b", "c"))
```
**`f1` could be renamed `match_prefix`. `f2` could be renamed `drop_last`. `f3` could be renamed `match_length` but there might be something better.

2. Take a function that you’ve written recently and spend 5 minutes brainstorming a better name for it and its arguments.

**I don't have anything right now.**

3. Compare and contrast `rnorm()` and `MASS::mvrnorm()`. How could you make them more consistent?

```{r}
?rnorm
?MASS::mvrnorm
```

**`rnorm` generates a random sample from the normal distribution, with parameters of `n` for sample size and `mean` and `sd` for the distribution. `mvrnorm` samples from the multivariate normal distribution, with parameters `n` for the number of samples required (default is 1) and vector `mu` for the means with a covariance matrix `Sigma` plus several other optional parameters. The presence/absence of a default sample size value and the SD vs covariance distinctions could cause confusion and might be made more consistent, but at the same time the options appear to be appropriate to each of the relevant procedures and distributions involved.**

4. Make a case for why `norm_r()`, `norm_d()` etc would be better than `rnorm()`, `dnorm()`. Make a case for the opposite.

**`norm_` as the prefix could make sense because you know which distribution you want to work with and can then bring up the list of functions to identify the task you want to perform using that distribution (random generation, density, etc.) On the other hand, beginning with the letter that identifies the task you want to perform, e.g., you know you need to generate a random sample so you type `r`, and then you're presented with a list of the available distributions to choose from - although frankly, a single letter will match a whole bunch of other functions and probably isn't going to be very helpful for autocompletion.**

# 19.4 Conditional execution

## 19.4.4 Exercises

1. What’s the difference between `if` and `ifelse()`? Carefully read the help and construct three examples that illustrate the key differences.

```{r}
?`if`
?ifelse
```

**`if` tests a single condition and then allows blocks of code to be executed depending on whether that condition was or, in combination with `else`, wasn't satisfied. On the other hand, `ifelse` can take a set of inputs to test (e.g., a vector of numbers) and will return an equal-length set with each element assigned its value depending on whether the test condition was or wasn't satisfied But all it can do is return those values rather than execute a sequence of operations.**

2. Write a greeting function that says “good morning”, “good afternoon”, or “good evening”, depending on the time of day. (Hint: use a time argument that defaults to `lubridate::now()`. That will make it easier to test your function.)

```{r}
say_gday <- function(time = now()) {
  time_hour <- hour(time)
  if (time_hour < 12) {
    greeting <- "morning"
  } else if (time_hour < 18) {
    greeting <- "afternoon"
  } else {
    greeting <- "evening"
  }
  writeLines(str_c("good ", greeting))
}
say_gday()
say_gday(dmy_hm("4/4/2017 09:30"))
say_gday(dmy_hm("4/4/2017 09:30pm"))
say_gday(dmy_hm("4/4/2017 16:20"))
```

3. Implement a `fizzbuzz` function. It takes a single number as input. If the number is divisible by three, it returns “fizz”. If it’s divisible by five it returns “buzz”. If it’s divisible by three and five, it returns “fizzbuzz”. Otherwise, it returns the number. Make sure you first write working code before you create the function.

```{r}
fizzbuzz <- function(x) {
  if ((x %% 3 == 0) && (x %% 5 == 0)) {
    return("fizzbuzz")
  } else if (x %% 3 == 0) {
    return("fizz")
  } else if (x %% 5 == 0) {
    return("buzz")
  } else {
    return(x)
  }
}
fizzbuzz(15)
fizzbuzz(9)
fizzbuzz(10)
fizzbuzz(8)
```

4. How could you use `cut()` to simplify this set of nested if-else statements?

```{r}
if (temp <= 0) {
  "freezing"
} else if (temp <= 10) {
  "cold"
} else if (temp <= 20) {
  "cool"
} else if (temp <= 30) {
  "warm"
} else {
  "hot"
}
```

```{r}
temps <- c(-8, 24, 35, 16, 7, 10)
cut(
  temps,
  breaks = c(-Inf, 0, 10, 20, 30, Inf),
  labels = c("freezing", "cold", "cool", "warm", "hot")
)
```

How would you change the call to `cut()` if I’d used `<` instead of `<=`? What is the other chief advantage of `cut()` for this problem? (Hint: what happens if you have many values in `temp`?)

```{r}
cut(
  temps,
  breaks = c(-Inf, 0, 10, 20, 30, Inf),
  labels = c("freezing", "cold", "cool", "warm", "hot"),
  right = FALSE # Makes the ranges closed on the left and open on the right
)
```

**`cut()` can deal with a vector filled with `temps` rather than needing to invoke a function with the `if` and `else` statements for every `temp` value.**

5. What happens if you use `switch()` with numeric values?

```{r}
switch(3, "duck(1)", "duck(2)", "goose") # option 3
switch(7 %% 3, "duck(1)", "duck(2)", "goose") # option 1
switch(7 / 3, "duck(1)", "duck(2)", "goose") # option 2
```

**It will choose the argument that corresponds in position to the number / numeric expression.**

6. What does this `switch()` call do? What happens if `x` is “e”?

```{r}
switch(x, 
  a = ,
  b = "ab",
  c = ,
  d = "cd"
)
```

Experiment, then carefully read the documentation.

```{r}
switch("e", 
  a = ,
  b = "ab",
  c = ,
  d = "cd"
)
switch("a", 
  a = ,
  b = "ab",
  c = ,
  d = "cd"
)
switch("b", 
  a = ,
  b = "ab",
  c = ,
  d = "cd"
)
switch("c", 
  a = ,
  b = "ab",
  c = ,
  d = "cd"
)
switch("d", 
  a = ,
  b = "ab",
  c = ,
  d = "cd"
)
```

**The help explains that when the expression is a character string then the string is matched to the names of elements, and if there is a match then that element is evaluated - unless it is missing, in which case it goes on to evaluate the next non-missing element. This explains why "a" and "c" each flow down to display the subsequent element's string. Because `"e"` doesn't match any element names, nothing is evaluated.**

# 19.5 Function arguments

## 19.5.5 Exercises

1. What does `commas(letters, collapse = "-")` do? Why?

```{r}
commas <- function(...) stringr::str_c(..., collapse = ", ")
commas(letters, collapse = "-")
```

**Because a `collapse` argument has been passed to `commas()` as part of its `...`, it is then passed through to `str_c` remaining as part of the `...`, colliding with rather than overriding the default `collapse` value.**

2. It’d be nice if you could supply multiple characters to the pad argument, e.g. `rule("Title", pad = "-+")`. Why doesn’t this currently work? How could you fix it?

```{r}
rule <- function(..., pad = "-") {
  title <- paste0(...)
  width <- getOption("width") - nchar(title) - 5
  cat(title, " ", stringr::str_dup(pad, width), "\n", sep = "")
}
rule("Important output")
rule("Title", pad = "-+")
```

**It doesn't work because the `pad` string is duplicated a number of times equal to the width that has been calculated, so if there are two characters it is twice as wide as needed.**

```{r}
rule2 <- function(..., pad = "-") {
  title <- paste0(...)
  width <- getOption("width") - nchar(title) - 5
  cat(title, " ", stringr::str_dup(pad, width / nchar(pad)), "\n", sep = "")
}
rule2("Important output")
rule2("Title", pad = "-+")
```

3. What does the `trim` argument to `mean()` do? When might you use it?

```{r}
?mean
```

**It specifies the proportion of cases at each end of the distribution that should be trimmed/dropped from the calculation. It would be useful if you have outliers that would have a strong effect on the mean and make it less representative of the central tendency in the observations, e.g., `trim = .025` would drop the most extreme values at each end and calculate the mean based on the remaining 95% of observations.**

4. The default value for the method argument to `cor()` is `c("pearson", "kendall", "spearman")`. What does that mean? What value is used by default?

```{r}
?cor
cor(1:10, 2:11, method = "wickham")
```

**By default it calculates the Pearson correlation. The vector specified as the default defines the acceptable values - if anything else is passed as an argument it will produce an error message.**
