---
title: "Chapter 20: Vectors"
output: html_notebook
---

```{r}
library(tidyverse)
```

# 20.3 Important types of atomic vector

## 20.3.5 Exercises

1. Describe the difference between `is.finite(x)` and `!is.infinite(x)`.

```{r}
?is.finite
```

**`is.finite` will return a `FALSE` element for not only `Inf` and `-Inf` but also for `NA` and `NaN` as they aren't finite values. On the other hand, `is.infinite` only returns `TRUE` for `Inf` and `-Inf`, so `!is.infinite` will return `FALSE` for those values but ends up returning `TRUE` for `NA` and `NaN`.**

```{r}
is.finite(c(0, Inf, -Inf, NA, NaN))
!is.infinite(c(0, Inf, -Inf, NA, NaN))
```

2. Read the source code for `dplyr::near()` (Hint: to see the source code, drop the `()`). How does it work?

```{r}
near
```

**It tests whether the absolute value of the difference between the two values is less than a tolerance threshold, which has a default value or can be overridden.**

3. A logical vector can take 3 possible values. How many possible values can an integer vector take? How many possible values can a double take? Use google to do some research.

```{r}
.Machine$integer.max
```

**This is `2 ^ 31 - 1` and is the maximum integer value R can handle. Negative values can go to the same absolute value. So the total number of possible values is above 4.2 billion, and then there are the special values such as `NA`.**

4. Brainstorm at least four functions that allow you to convert a double to an integer. How do they differ? Be precise.

**There is an `as.integer()` function that will coerce values to be integers. The rounding functions (`round`, `floor`, etc.) appear to return doubles even though the value has been made a whole number.**

5. What functions from the readr package allow you to turn a string into logical, integer, and double vector?

```{r}
?parse_logical
```

**The `parse_` functions take a character vector and parse them to other vector types.**

# 20.4 Using atomic vectors

## 20.4.6 Exercises

1. What does `mean(is.na(x))` tell you about a vector `x`? What about `sum(!is.finite(x)`)?

**`mean(is.na(x))` tells you the proportion of elements in the vector that are missing values. `sum(!is.finite(x))` tells you how many of the elements in the vector aren't finite values.**

2. Carefully read the documentation of `is.vector()`. What does it actually test for? Why does `is.atomic()` not agree with the definition of atomic vectors above?

```{r}
?is.vector
?is.atomic
```

**`is.vector()` tests whether `x` is a vector of the specified mode (default is "any") that has no attributes other than names. `is.atomic()` returns `TRUE` not only if `x` is of an atomic type, but also for `NULL`.**

3. Compare and contrast `setNames()` with `purrr::set_names()`.

```{r}
?setNames
?purrr::set_names
```

**`set_names` adjusts the defaults so that if a vector is passed in without a second argument, the vector will be named with itself. As a tidyverse function it will also include checks to prevent silent recycling.**

4. Create functions that take a vector as input and returns:

  1. The last value. Should you use `[` or `[[`?
  2. The elements at even numbered positions.
  3. Every element except the last value.
  4. Only even numbers (and no missing values).

```{r}
get_last <- function(x) {
  x[[length(x)]] # use [[ because we want just the single value
}

get_even_pos <- function(x) {
  x[c(FALSE, TRUE)] # The two elements will be recycled to the full length of x
}

drop_last <- function(x) {
  x[-length(x)]
}

get_even_val <- function(x) {
  x[!is.na(x) & x %% 2 == 0]
}

test <- c(NA, 0:5, NA, 6:10)
get_last(test)
get_even_pos(test)
drop_last(test)
get_even_val(test)
```

5. Why is `x[-which(x > 0)]` not the same as `x[x <= 0]`?

```{r}
test2 <- c(-10, 8, NA, 0)
test2[-which(test2 > 0)]
test2[test2 <= 0]
```

**I'll leave this for now.**

6. What happens when you subset with a positive integer that’s bigger than the length of the vector? What happens when you subset with a name that doesn’t exist?

```{r}
test3 <- c(a = 1, b = 2, c = 3)
test3[2:4]
test3[c("c", "d")]
```

**Subsetting with too large a number returns `NA` for any elements that are beyond the length of the original vector. Using a name that doesn't exist will also return an `NA`.**
