---
title: "Chapter 21: Iteration"
output: html_notebook
---

```{r}
library(tidyverse)
```

# 21.2 For loops

## 21.2.1 Exercises

1. Write for loops to:

  1. Compute the mean of every column in `mtcars`.
  2. Determine the type of each column in `nycflights13::flights`.
  3. Compute the number of unique values in each column of `iris`.
  4. Generate 10 random normals for each of $\mu = -10$, $0$, $10$, and $100$.

```{r}
for_out_1 <- vector("double", ncol(mtcars))
for (i in seq_along(mtcars)) {
  for_out_1[[i]] <- mean(mtcars[[i]], na.rm = TRUE)
}
for_out_1
```

```{r}
library(nycflights13)
for_out_2 <- vector("character", ncol(flights))
for (i in seq_along(flights)) {
  for_out_2[[i]] <- typeof(flights[[i]])
}
for_out_2
```

```{r}
for_out_3 <- vector("integer", ncol(iris))
for (i in seq_along(iris)) {
  for_out_3[[i]] <- length(unique(iris[[i]]))
}
for_out_3
```

```{r}
means <- c(-10, 0, 10, 100)
for_out_4 <- vector("list", length(means))
names(for_out_4) <- means
for(i in seq_along(for_out_4)) {
  for_out_4[[i]] <- rnorm(10, mean = means[[i]])
}
for_out_4
```

Think about the output, sequence, and body before you start writing the loop.

2. Eliminate the for loop in each of the following examples by taking advantage of an existing function that works with vectors:

```{r}
out <- ""
for (x in letters) {
  out <- stringr::str_c(out, x)
}
out
```

```{r}
library(stringr)
out2 <- str_c(letters, collapse = "")
out2
```

```{r}
x <- sample(100)
sd <- 0
for (i in seq_along(x)) {
  sd <- sd + (x[i] - mean(x)) ^ 2
}
sd <- sqrt(sd / (length(x) - 1))
sd
```

```{r}
# Use the same sample above - ensure this block is run after above block
sd2 <- sqrt(sum((x - mean(x)) ^ 2) / (length(x) - 1))
sd2
```

```{r}
x <- runif(100)
out <- vector("numeric", length(x))
out[1] <- x[1]
for (i in 2:length(x)) {
  out[i] <- out[i - 1] + x[i]
}
out
```

```{r}
# The transform chapter referred to a function to calculate cmulative sums
cumsum(x)
```

3. Combine your function writing and for loop skills:

  1. Write a for loop that `prints()` the lyrics to the children’s song “Alice the camel”.
  2. Convert the nursery rhyme “ten in the bed” to a function. Generalise it to any number of people in any sleeping structure.
  3. Convert the song “99 bottles of beer on the wall” to a function. Generalise to any number of any vessel containing any liquid on any surface.

**I might come back to these later.**

4. It’s common to see for loops that don’t preallocate the output and instead increase the length of a vector at each step:

```{r}
output <- vector("integer", 0)
for (i in seq_along(x)) {
  output <- c(output, lengths(x[[i]]))
}
output
```

How does this affect performance? Design and execute an experiment.

```{r}
library(microbenchmark)

input <- rnorm(100)

no_prealloc <- function(y) {
  output <- vector("integer", 0)
  for (i in seq_along(y)) {
    output <- c(output, lengths(y[[i]]))
  }
  output
}

prealloc <- function(y) {
  output <- vector("integer", length(y))
  for(i in seq_along(y)) {
    output <- c(output, lengths(y[[i]]))
  }
  output
}

microbenchmark(
  no_prealloc(input),
  prealloc(input),
  times = 10000
)
```

**I'm consistently getting results where the preallocated function tends to be slower on average. They're identical functions except for the vector allocation, so I can't see a clear explanation and don't have another computer to test with at the moment.**

```{r}
input2 <- runif(5000)

alt_no_pre <- function(x) {
  out <- vector("numeric", 0)
  out[1] <- x[1]
  for (i in 2:length(x)) {
    out[i] <- out[i - 1] + x[i]
  }
  out
}

alt_pre <- function(x) {
  out <- vector("numeric", length(x))
  out[1] <- x[1]
  for (i in 2:length(x)) {
    out[i] <- out[i - 1] + x[i]
  }
  out
}

microbenchmark(
  alt_no_pre(input2),
  alt_pre(input2)
)

```

**I did some more testing with this second pair of functions. It seems that the size of the vector being created through the for loop (ie, number of iterations) has a big effect - at 5000 the mean time without preallocation is 10x that with preallocation, but at 1000 it was 2x. But changing the number of iterations involved didn't seem to produce the same sort of outcome. I'm still a bit confused.**
