---
title: "Chapter 21: Iteration"
output: html_notebook
---

```{r}
library(tidyverse)
```

# 21.2 For loops

## 21.2.1 Exercises

1. Write for loops to:

  1. Compute the mean of every column in `mtcars`.
  2. Determine the type of each column in `nycflights13::flights`.
  3. Compute the number of unique values in each column of `iris`.
  4. Generate 10 random normals for each of $\mu = -10$, $0$, $10$, and $100$.

```{r}
for_out_1 <- vector("double", ncol(mtcars))
for (i in seq_along(mtcars)) {
  for_out_1[[i]] <- mean(mtcars[[i]], na.rm = TRUE)
}
for_out_1
```

```{r}
library(nycflights13)
for_out_2 <- vector("character", ncol(flights))
for (i in seq_along(flights)) {
  for_out_2[[i]] <- typeof(flights[[i]])
}
for_out_2
```

```{r}
for_out_3 <- vector("integer", ncol(iris))
for (i in seq_along(iris)) {
  for_out_3[[i]] <- length(unique(iris[[i]]))
}
for_out_3
```

```{r}
means <- c(-10, 0, 10, 100)
for_out_4 <- vector("list", length(means))
names(for_out_4) <- means
for(i in seq_along(for_out_4)) {
  for_out_4[[i]] <- rnorm(10, mean = means[[i]])
}
for_out_4
```

Think about the output, sequence, and body before you start writing the loop.

2. Eliminate the for loop in each of the following examples by taking advantage of an existing function that works with vectors:

```{r}
out <- ""
for (x in letters) {
  out <- stringr::str_c(out, x)
}
out
```

```{r}
library(stringr)
out2 <- str_c(letters, collapse = "")
out2
```

```{r}
x <- sample(100)
sd <- 0
for (i in seq_along(x)) {
  sd <- sd + (x[i] - mean(x)) ^ 2
}
sd <- sqrt(sd / (length(x) - 1))
sd
```

```{r}
# Use the same sample above - ensure this block is run after above block
sd2 <- sqrt(sum((x - mean(x)) ^ 2) / (length(x) - 1))
sd2
```

```{r}
x <- runif(100)
out <- vector("numeric", length(x))
out[1] <- x[1]
for (i in 2:length(x)) {
  out[i] <- out[i - 1] + x[i]
}
out
```

```{r}
# The transform chapter referred to a function to calculate cmulative sums
cumsum(x)
```

3. Combine your function writing and for loop skills:

  1. Write a for loop that `prints()` the lyrics to the children’s song “Alice the camel”.
  2. Convert the nursery rhyme “ten in the bed” to a function. Generalise it to any number of people in any sleeping structure.
  3. Convert the song “99 bottles of beer on the wall” to a function. Generalise to any number of any vessel containing any liquid on any surface.

**I might come back to these later.**

4. It’s common to see for loops that don’t preallocate the output and instead increase the length of a vector at each step:

```{r}
output <- vector("integer", 0)
for (i in seq_along(x)) {
  output <- c(output, lengths(x[[i]]))
}
output
```

How does this affect performance? Design and execute an experiment.

```{r}
library(microbenchmark)

input <- rnorm(100)

no_prealloc <- function(y) {
  output <- vector("integer", 0)
  for (i in seq_along(y)) {
    output <- c(output, lengths(y[[i]]))
  }
  output
}

prealloc <- function(y) {
  output <- vector("integer", length(y))
  for(i in seq_along(y)) {
    output <- c(output, lengths(y[[i]]))
  }
  output
}

microbenchmark(
  no_prealloc(input),
  prealloc(input),
  times = 10000
)
```

**I'm consistently getting results where the preallocated function tends to be slower on average. They're identical functions except for the vector allocation, so I can't see a clear explanation and don't have another computer to test with at the moment.**

```{r}
input2 <- runif(5000)

alt_no_pre <- function(x) {
  out <- vector("numeric", 0)
  out[1] <- x[1]
  for (i in 2:length(x)) {
    out[i] <- out[i - 1] + x[i]
  }
  out
}

alt_pre <- function(x) {
  out <- vector("numeric", length(x))
  out[1] <- x[1]
  for (i in 2:length(x)) {
    out[i] <- out[i - 1] + x[i]
  }
  out
}

microbenchmark(
  alt_no_pre(input2),
  alt_pre(input2)
)

```

**I did some more testing with this second pair of functions. It seems that the size of the vector being created through the for loop (ie, number of iterations) has a big effect - at 5000 the mean time without preallocation is 10x that with preallocation, but at 1000 it was 2x. But changing the number of iterations involved didn't seem to produce the same sort of outcome. I'm still a bit confused.**

# 21.3 For loop variations

## 21.3.5 Exercises

1. Imagine you have a directory full of CSV files that you want to read in. You have their paths in a vector, `files <- dir("data/", pattern = "\\.csv$", full.names = TRUE)`, and now want to read each one with `read_csv()`. Write the for loop that will load them into a single data frame.

```
# Use a list and make it a df at the end for efficiency
df <- vector("list", length(files))
for(i in seq_along(files)) {
  df[[i]] <- read_csv(files[[i]])
}
df <- bind_rows(df)
```

2. What happens if you use `for (nm in names(x))` and `x` has no names? What if only some of the elements are named? What if the names are not unique?

**I might come back to writing tests for these.**

3. Write a function that prints the mean of each numeric column in a data frame, along with its name. For example, `show_mean(iris)` would print:

```
show_mean(iris)
#> Sepal.Length: 5.84
#> Sepal.Width:  3.06
#> Petal.Length: 3.76
#> Petal.Width:  1.20
```

```{r}
print_col_means <- function(df) {
  # No vector preallocated because we won't store anything - we can print directly from for loop
  for(column in names(df)) {
    if (is_double(df[[column]])) {
      writeLines((str_c(column, ": ", format(mean(df[[column]], na.rm = TRUE), digits = 2, nsmall = 2))))
    }
  }
}
print_col_means(iris)
```

```{r}
print_col_means(mtcars)
```

(Extra challenge: what function did I use to make sure that the numbers lined up nicely, even though the variable names had different lengths?)

**I might come back to this.**

4. What does this code do? How does it work?

```{r}
mtcars
trans <- list( 
  disp = function(x) x * 0.0163871,
  am = function(x) {
    factor(x, labels = c("auto", "manual"))
  }
)
for (var in names(trans)) {
  mtcars[[var]] <- trans[[var]](mtcars[[var]])
}
mtcars
```

**It transforms two variables in the `mtcars` data - `disp` is replaced by its original value multiplied by 0.0163871 (which searching suggests is the process to convert from cubic inches to litres), and `am` is converted from a numeric variable to a factor with `auto` and `manual` labels. It works by first creating a list thatcontains two elements, each of which is a function that definesthe necessary transformations. Then a for loopiterates across the names of those list elements (which correspond to the variable names in the dataset) and assigns to those variables the corresponding list element - which passes the original value in as an argument to be used within the function that is defined in the list element.**

# 21.4 For loops vs. functionals

## 21.4.1 Exercises

1. Read the documentation for `apply()`. In the 2d case, what two for loops does it generalise?

```{r}
?apply
```

**It can apply functions to subscripts along the rows or the columns, so it appears to generalise `for (i in seq_along(xs))` as we've used it to iterate across columns. I'm not sure how to define the second loop type.**

2. Adapt `col_summary()` so that it only applies to numeric columns. You might want to start with an `is_numeric()` function that returns a logical vector that has a `TRUE` corresponding to each numeric column.

```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = letters[1:10]
)

col_summary <- function(df, fun) {
  isnum <- vector("logical", length(df))
  for (i in seq_along(df)) {
    isnum[[i]] = is_numeric(df[[i]])
  }
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    if (isnum[i]) {
      out[i] <- fun(df[[i]])
    } else {
      out[i] <- NA
    }
  }
  out
}

col_summary(df, mean)
```

# 21.5 The map functions

## 21.5.3 Exercises

1. Write code that uses one of the `map` functions to:

  1. Compute the mean of every column in `mtcars`.
  2. Determine the type of each column in `nycflights13::flights`.
  3. Compute the number of unique values in each column of `iris`.
  4. Generate 10 random normals for each of $\mu=−10$, $0$, $10$, and $100$.

```{r}
# NB: Restart session before running this chunk or the transformations to `mtcars` above will cause an error on the `am` column
library(tidyverse)
library(nycflights13)
map_dbl(mtcars, mean)
```

```{r}
map_chr(flights, typeof)
```

```{r}
map_int(iris, ~length(unique(.)))
```

```{r}
means <- c("-10" = -10, "0" = 0, "10" = 10, "100" = 100)
map(means, rnorm, n = 10)
```

2. How can you create a single vector that for each column in a data frame indicates whether or not it’s a factor?

```{r}
map_lgl(iris, is.factor)
```

3. What happens when you use the map functions on vectors that aren’t lists? What does `map(1:5, runif)` do? Why?

```{r}
map(1:5, runif)
```

**It applies the mapped function to each element of the vector being passed in. In this case, `runif` runs to produce samples from size 1 up to size 5.**

4. What does `map(-2:2, rnorm, n = 5)` do? Why? What does `map_dbl(-2:2, rnorm, n = 5)` do? Why?

```{r}
map(-2:2, rnorm, n = 5)
```

**It produces five random normal samples, each of size 5, but with the population mean for each sampling process being `-2`, `-1`, `0`, `1`, and `2`. The help for `rnorm` indicates that its function definition is `rnorm(n, mean = 0, sd = 1)` - it appears that because an `n` argument is passed through the `map` function, the vector being passed is used in the place of the next parameter which is `mean`.**

```{r}
map_dbl(-2:2, rnorm, n = 5)
```

**It produces an error because `map_dbl` expects a single value (length 1 atomic vector) to be produced by each function call, rather than a vector of multiple (5) numbers.**

5. Rewrite `map(x, function(df) lm(mpg ~ wt, data = df))` to eliminate the anonymous function.

```{r}
map(mtcars, ~lm(mpg ~ wt, data = .))
```

**This appears not to work because `map` is being passed a single data frame rather than something to iterate over - compare with the example in this section where `mtcars` was passed through a `split` function call, which produced a list containing three data frames. We can create a single-item list containing our data frame, which will then be passed to the mapped function.**

```{r}
list(mtcars) %>%
  map(~lm(mpg ~ wt, data = .))
```

# 21.9 Other patterns of for loops

## 21.9.3 Exercises

1. Implement your own version of `every()` using a for loop. Compare it with `purrr::every()`. What does purrr’s version do that your version doesn’t?

```{r}
every_for <- function(x, predicate) {
  for (i in seq_along(x)) {
    if(!predicate(x[[i]])) {
      return(FALSE)
    }
  }
  return(TRUE)
}
every1 <- list(1:5, letters, list(10))
every_for(every1, is_vector)
every_for(every1, is_numeric)
```

```{r}
every
```

**The major differences appear to be that my function doesn't use `as_function()`, which allows more complexity in what is passed through and used within the test function, and my functuion doesn't handle additional arguments through `...`. **

2. Create an enhanced `col_sum()` that applies a summary function to every numeric column in a data frame.

```{r}
col_sum2 <- function(df, fn, ...) {
  fn <- as_function(fn, ...)
  map(keep(df, is_double), fn, ...)
}
col_sum2(mtcars, mean)
col_sum2(iris, mean)
```

3. A possible base R equivalent of `col_sum()` is:

```{r}
col_sum3 <- function(df, f) {
  is_num <- sapply(df, is.numeric)
  df_num <- df[, is_num]

  sapply(df_num, f)
}
```

But it has a number of bugs as illustrated with the following inputs:

```{r}
df <- tibble(
  x = 1:3, 
  y = 3:1,
  z = c("a", "b", "c")
)
# OK
col_sum3(df, mean)
# Has problems: don't always return numeric vector
col_sum3(df[1:2], mean)
col_sum3(df[1], mean)
col_sum3(df[0], mean)
```

What causes the bugs?

**If the `sapply` function is passed out of bounds index values then it will return an empty list that isn't of a numeric type, which then causes the next subsetting line to fail. One way to overcome the error is to coerce the `sapply` output to a numeric value, although this ends up just returning an empty list when the summary function is applied.**

```{r}
col_sum4 <- function(df, f) {
  is_num <- as.numeric(sapply(df, is.numeric))
  df_num <- df[, is_num]

  sapply(df_num, f)
}

col_sum4(df[1:2], mean)
col_sum4(df[1], mean)
col_sum4(df[0], mean)
```

