---
title: "Chapter 21: Iteration"
output: html_notebook
---

```{r}
library(tidyverse)
```

# 21.2 For loops

## 21.2.1 Exercises

1. Write for loops to:

  1. Compute the mean of every column in `mtcars`.
  2. Determine the type of each column in `nycflights13::flights`.
  3. Compute the number of unique values in each column of `iris`.
  4. Generate 10 random normals for each of $\mu = -10$, $0$, $10$, and $100$.

```{r}
for_out_1 <- vector("double", ncol(mtcars))
for (i in seq_along(mtcars)) {
  for_out_1[[i]] <- mean(mtcars[[i]], na.rm = TRUE)
}
for_out_1
```

```{r}
library(nycflights13)
for_out_2 <- vector("character", ncol(flights))
for (i in seq_along(flights)) {
  for_out_2[[i]] <- typeof(flights[[i]])
}
for_out_2
```

```{r}
for_out_3 <- vector("integer", ncol(iris))
for (i in seq_along(iris)) {
  for_out_3[[i]] <- length(unique(iris[[i]]))
}
for_out_3
```

```{r}
means <- c(-10, 0, 10, 100)
for_out_4 <- vector("list", length(means))
names(for_out_4) <- means
for(i in seq_along(for_out_4)) {
  for_out_4[[i]] <- rnorm(10, mean = means[[i]])
}
for_out_4
```

Think about the output, sequence, and body before you start writing the loop.

2. Eliminate the for loop in each of the following examples by taking advantage of an existing function that works with vectors:

```{r}
out <- ""
for (x in letters) {
  out <- stringr::str_c(out, x)
}
out
```

```{r}
library(stringr)
out2 <- str_c(letters, collapse = "")
out2
```

```{r}
x <- sample(100)
sd <- 0
for (i in seq_along(x)) {
  sd <- sd + (x[i] - mean(x)) ^ 2
}
sd <- sqrt(sd / (length(x) - 1))
sd
```

```{r}
# Use the same sample above - ensure this block is run after above block
sd2 <- sqrt(sum((x - mean(x)) ^ 2) / (length(x) - 1))
sd2
```

```{r}
x <- runif(100)
out <- vector("numeric", length(x))
out[1] <- x[1]
for (i in 2:length(x)) {
  out[i] <- out[i - 1] + x[i]
}
out
```

```{r}
# The transform chapter referred to a function to calculate cmulative sums
cumsum(x)
```

3. Combine your function writing and for loop skills:

  1. Write a for loop that `prints()` the lyrics to the children’s song “Alice the camel”.
  2. Convert the nursery rhyme “ten in the bed” to a function. Generalise it to any number of people in any sleeping structure.
  3. Convert the song “99 bottles of beer on the wall” to a function. Generalise to any number of any vessel containing any liquid on any surface.

**I might come back to these later.**

4. It’s common to see for loops that don’t preallocate the output and instead increase the length of a vector at each step:

```{r}
output <- vector("integer", 0)
for (i in seq_along(x)) {
  output <- c(output, lengths(x[[i]]))
}
output
```

How does this affect performance? Design and execute an experiment.

```{r}
library(microbenchmark)

input <- rnorm(100)

no_prealloc <- function(y) {
  output <- vector("integer", 0)
  for (i in seq_along(y)) {
    output <- c(output, lengths(y[[i]]))
  }
  output
}

prealloc <- function(y) {
  output <- vector("integer", length(y))
  for(i in seq_along(y)) {
    output <- c(output, lengths(y[[i]]))
  }
  output
}

microbenchmark(
  no_prealloc(input),
  prealloc(input),
  times = 10000
)
```

**I'm consistently getting results where the preallocated function tends to be slower on average. They're identical functions except for the vector allocation, so I can't see a clear explanation and don't have another computer to test with at the moment.**

```{r}
input2 <- runif(5000)

alt_no_pre <- function(x) {
  out <- vector("numeric", 0)
  out[1] <- x[1]
  for (i in 2:length(x)) {
    out[i] <- out[i - 1] + x[i]
  }
  out
}

alt_pre <- function(x) {
  out <- vector("numeric", length(x))
  out[1] <- x[1]
  for (i in 2:length(x)) {
    out[i] <- out[i - 1] + x[i]
  }
  out
}

microbenchmark(
  alt_no_pre(input2),
  alt_pre(input2)
)

```

**I did some more testing with this second pair of functions. It seems that the size of the vector being created through the for loop (ie, number of iterations) has a big effect - at 5000 the mean time without preallocation is 10x that with preallocation, but at 1000 it was 2x. But changing the number of iterations involved didn't seem to produce the same sort of outcome. I'm still a bit confused.**

# 21.3 For loop variations

## 21.3.5 Exercises

1. Imagine you have a directory full of CSV files that you want to read in. You have their paths in a vector, `files <- dir("data/", pattern = "\\.csv$", full.names = TRUE)`, and now want to read each one with `read_csv()`. Write the for loop that will load them into a single data frame.

```
# Use a list and make it a df at the end for efficiency
df <- vector("list", length(files))
for(i in seq_along(files)) {
  df[[i]] <- read_csv(files[[i]])
}
df <- bind_rows(df)
```

2. What happens if you use `for (nm in names(x))` and `x` has no names? What if only some of the elements are named? What if the names are not unique?

**I might come back to writing tests for these.**

3. Write a function that prints the mean of each numeric column in a data frame, along with its name. For example, `show_mean(iris)` would print:

```
show_mean(iris)
#> Sepal.Length: 5.84
#> Sepal.Width:  3.06
#> Petal.Length: 3.76
#> Petal.Width:  1.20
```

```{r}
print_col_means <- function(df) {
  # No vector preallocated because we won't store anything - we can print directly from for loop
  for(column in names(df)) {
    if (is_double(df[[column]])) {
      writeLines((str_c(column, ": ", format(mean(df[[column]], na.rm = TRUE), digits = 2, nsmall = 2))))
    }
  }
}
print_col_means(iris)
```

```{r}
print_col_means(mtcars)
```

(Extra challenge: what function did I use to make sure that the numbers lined up nicely, even though the variable names had different lengths?)

**I might come back to this.**

4. What does this code do? How does it work?

```{r}
mtcars
trans <- list( 
  disp = function(x) x * 0.0163871,
  am = function(x) {
    factor(x, labels = c("auto", "manual"))
  }
)
for (var in names(trans)) {
  mtcars[[var]] <- trans[[var]](mtcars[[var]])
}
mtcars
```

**It transforms two variables in the `mtcars` data - `disp` is replaced by its original value multiplied by 0.0163871 (which searching suggests is the process to convert from cubic inches to litres), and `am` is converted from a numeric variable to a factor with `auto` and `manual` labels. It works by first creating a list thatcontains two elements, each of which is a function that definesthe necessary transformations. Then a for loopiterates across the names of those list elements (which correspond to the variable names in the dataset) and assigns to those variables the corresponding list element - which passes the original value in as an argument to be used within the function that is defined in the list element.**
